---
title: "Reds Hackathon Code"
author: "Nathan Backman, Nolan Pittman, Jake Balek, Hunter Geise, Danielle Napierski"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# libraries #

```{r}
library(tidyverse)
library(caret)
library(rsample)
library(cluster)    
library(factoextra) 
library(dendextend)
library(NbClust)
library(uuid)
library(baseballr)
```

# read in data
```{r}
savant <- read.csv("savant_pitch_level.csv")
fangraphs <- read.csv("fangraphs_season_level.csv")
```

# add statcast zones #

# functions #

```{r}
heart_z <- function(sz_bot,sz_top) {
  mid <- (sz_bot + sz_top)/2
  bot <- mid - ((mid - sz_bot)*.67)
  top <- mid + ((sz_top - mid)*.67)
  return(c(bot,top))
}

shadow_z <- function(sz_bot,sz_top) {
  mid <- (sz_bot + sz_top)/2
  bot <- sz_bot - (mid*.33)
  top <- sz_top + (mid*.33)
  return(c(bot,top))
}

chase_z <- function(sz_bot,sz_top) {
  mid <- (sz_bot + sz_top)/2
  bot <- sz_bot - (mid*2)
  top <- sz_top + (mid*2)
  return(c(bot,top))
}
```

# variables #

```{r}
heart_x <- 6.7/12
shadow_x <- 13.3/12
chase_x <- 20/12

pb <- txtProgressBar(min = 0,      
                     max = nrow(data), 
                     style = 3,  
                     width = 50, 
                     char = "=")
```

# loop #

```{r}

savant %>% 
  mutate(heart = 0,
         shadow = 0,
         chase = 0,
         waste = 0) -> savant

for (row in 1:nrow(savant)) {
  if (is.na(savant$plate_x[row]) == T |
      is.na(savant$plate_z[row]) == T |
      is.na(savant$sz_top[row]) == T |
      is.na(savant$sz_bot[row]) == T) {
    savant$heart[row] <- NA
    savant$shadow[row] <- NA
    savant$chase[row] <- NA
    savant$waste[row] <- NA
  } else if (na.exclude(savant$plate_x[row]) < heart_x &
             na.exclude(savant$plate_x[row]) > -heart_x &
             savant$plate_z[row] <
             heart_z(savant$sz_bot[row],savant$sz_top[row])[2] &
             savant$plate_z[row] >
             heart_z(savant$sz_bot[row],savant$sz_top[row])[1]) {
    savant$heart[row] <- 1
  } else if (savant$plate_x[row] < shadow_x &
             savant$plate_x[row] > -shadow_x &
             savant$plate_z[row] <
             shadow_z(savant$sz_bot[row],savant$sz_top[row])[2] &
             savant$plate_z[row] >
             shadow_z(savant$sz_bot[row],savant$sz_top[row])[1]) {
    savant$shadow[row] <- 1
  } else if (savant$plate_x[row] < chase_x &
             savant$plate_x[row] > -chase_x &
             savant$plate_z[row] <
             chase_z(savant$sz_bot[row],savant$sz_top[row])[2] &
             savant$plate_z[row] >
             chase_z(savant$sz_bot[row],savant$sz_top[row])[1]) {
    savant$chase[row] <- 1
  } else {
    savant$waste[row] <- 1
  }
  setTxtProgressBar(pb, row)
}


  if (row %% 15044 == 0) {
    print(glue("{round(row/nrow(savant)*100,0)}% done."))
  } else {
    invisible()
  }
```

# get game situation #

```{r}
savant %>%
  ungroup() %>% 
  mutate(
    on_1b_dummy = ifelse(is.na(on_1b), 0, 1),
    on_2b_dummy = ifelse(is.na(on_2b), 0, 1),
    on_3b_dummy = ifelse(is.na(on_3b), 0, 1)
  ) -> savant

paste(savant$on_1b_dummy,                  
      savant$on_2b_dummy,
      savant$on_3b_dummy,
      savant$outs_when_up, 
      sep = "-") -> savant$on1b_on2B_on3b_outs

team_mappings <- c("LAA" = "108", "AZ" = "109", "BAL" = "110", 
                   "BOS" = "111", "CHC" = "112", "CIN" = "113", 
                   "CLE" = "114", "COL" = "115", "DET" = "116", 
                   "HOU" = "117", "KC" = "118", "LAD" = "119", 
                   "WSH" = "120", "OAK" = "133", "PIT" = "134",
                   "SD" = "135", "SEA" = "136", "SF" = "137" ,
                   "STL" = "138", "TB" = "139", "TEX" = "140", 
                   "TOR" = "141", "MIN" = "142", "PHI" = "143", 
                   "ATL" = "144", "CWS" = "145", "MIA" = "146", 
                   "NYY" = "147", "NYM" = "121", "MIL" = "158" )

savant %>% 
  ungroup() %>% 
  group_by(game_pk, pitcher, at_bat_number) %>% 
  arrange(game_pk, pitcher, pitch_number_appearance) %>% 
  mutate(ab_id = UUIDgenerate()) %>% 
  filter(pitch_number_appearance == 1) %>% 
  mutate(game_start = ifelse(balls == 0 &
                               strikes == 0 &
                               at_bat_number == 1,
                             1, 0)) %>% 
  filter(game_start == 0) %>% 
  ungroup() -> savant2

savant2 %>% 
  select(game_pk, ab_id, pitcher, game_year, game_date, game_situation, 
         inning, inning_topbot, home_team, away_team) %>% 
  mutate(home_team_id = ifelse(home_team %in% names(team_mappings),
                               team_mappings[home_team], home_team),
         away_team_id = ifelse(away_team %in% names(team_mappings),
                               team_mappings[away_team], away_team),
         team_id = ifelse(inning_topbot == "Top", 
                          home_team_id,
                          away_team_id)) -> pitcher_management

map(pitcher_management$ab_id, function(x) {
  savant2 %>% 
    ungroup() %>%
    filter(ab_id == x) %>% 
    select(game_pk) %>% 
    distinct() %>% 
    .[[1]] -> game_pk
  savant2 %>% 
    ungroup() %>%
    filter(ab_id == x) %>% 
    select(pitcher) %>% 
    distinct() %>% 
    .[[1]] -> pitcher
  savant2 %>% 
    ungroup() %>%
    filter(ab_id == x) %>% 
    select(inning) %>% 
    distinct() %>% 
    .[[1]] -> inning
  savant2 %>% 
    ungroup() %>%
    filter(game_pk == game_pk &
             pitcher == pitcher &
             inning == inning) %>% 
    drop_na(delta_run_exp) %>% 
    summarise(mean_delta_run_exp = mean(delta_run_exp))
}, .progress = TRUE) %>% 
  bind_rows() -> d_run_exp

cbind(pitcher_management, d_run_exp) -> pitcher_management
```

# manipulate fangraphs data with statcast measurements #

```{r}
savant %>%
  ungroup() %>% 
  drop_na(launch_speed_angle) %>% 
  mutate(
    weak = ifelse(launch_speed_angle == "1", 1, 0),
    topped = ifelse(launch_speed_angle == "2", 1, 0),
    under = ifelse(launch_speed_angle == "3", 1, 0),
    flare = ifelse(launch_speed_angle == "4", 1, 0),
    solid = ifelse(launch_speed_angle == "5", 1, 0),
    barrel = ifelse(launch_speed_angle == "6", 1, 0),
  ) %>% 
  group_by(pitcher, game_year, role_key) %>% 
  summarise(
    weak_pct = sum(weak)/n(),
    topped_pct = sum(topped)/n(),
    under_pct = sum(under)/n(),
    flare_pct = sum(flare)/n(),
    solid_pct = sum(solid)/n(),
    barrel_pct = sum(barrel)/n()
  ) -> statcast_bb

left_join(fangraphs, statcast_bb,
          by = c("MLBAMID" = "pitcher",
                 "Season" = "game_year",
                 "Role" = "role_key")) -> fangraphs

savant %>% 
  filter(bb_type != "") %>% 
  mutate(
    gb = ifelse(bb_type == "ground_ball", 1, 0),
    fb = ifelse(bb_type == "fly_ball", 1, 0),
    ld = ifelse(bb_type == "line_drive", 1, 0),
    pu = ifelse(bb_type == "popup", 1, 0)
  ) %>% 
  group_by(pitcher, game_year, role_key) %>% 
  summarise(
    gb_pct = sum(gb)/n(),
    fb_pct = sum(fb)/n(),
    ld_pct = sum(ld)/n(),
    pu_pct = sum(pu)/n()
  ) -> statcast_bb_2

left_join(fangraphs, statcast_bb_2,
          by = c("MLBAMID" = "pitcher",
                 "Season" = "game_year",
                 "Role" = "role_key")) -> fangraphs

savant %>% 
  drop_na(shadow) %>% 
  mutate(shadow_pitch = ifelse(shadow == "1", 1, 0)) %>% 
  group_by(pitcher, game_year, role_key) %>% 
  summarise(shadow_rate = sum(shadow_pitch)/n()) -> shadow_rate

left_join(fangraphs, shadow_rate,
          by = c("MLBAMID" = "pitcher",
                 "Season" = "game_year",
                 "Role" = "role_key")) -> fangraphs
```

# creating clusters based on batted balls #

## Data Manipulation

```{r}
predictors <- c("weak_pct", "topped_pct", "under_pct", "flare_pct",
                "solid_pct", "barrel_pct",
                "gb_rate", "fb_rate", "ld_rate", "pu_rate")

# reference data
fangraphs %>%
  group_by(Season) %>%
  filter(TBF >= 50) %>%
  mutate(across(.cols = predictors, ~ntile(., 100), 
                .names = "{.col}_percentile")) %>%
  ungroup() %>% 
  select(MLBAMID, Season, Role, all_of(predictors),
         contains("percentile")) -> agnes_data 

# model data
agnes_data %>% 
  select(contains("percentile")) -> model_data 
```

## Optimizing K Clusters

```{r}
set.seed(3422354)

# WSS method
fviz_nbclust(model_data, FUN = hcut, method = "wss")$data %>% 
  rename(wss = y) -> wss

# Silhouette method
fviz_nbclust(model_data, FUN = hcut, 
             method = "silhouette")$data %>% 
  rename(mean_silhouette_width = y) -> silhouette

# Gap stat method
clusGap(model_data, FUN = hcut, K.max = 10,
        B = 50, method = "firstSEmax") -> gap_stat

gap_stat[["Tab"]] %>% 
  as_tibble() %>% 
  mutate(clusters = row_number()) -> gap_stat_table

# Visualize gap stat
fviz_gap_stat(gap_stat)

# look at all metrics
merge(
  merge(gap_stat_table, wss, by = "clusters"), 
  silhouette, by = "clusters"
  ) -> all_metrics
```

## Final Clustering

```{r}
set.seed(123)

# final clusters
final <- agnes(model_data, method = "ward", metric = "euclidean")
#use single with normal data

# optimal bb clusters
k <- 4 
final_clusters <- cutree(as.hclust(final), k = k)
table(final_clusters)

# Visualizing clusters
fviz_cluster(list(data = model_data, 
                  cluster = final_clusters), labelsize = 4)

agnes_data %>% 
  mutate(cluster = final_clusters) -> bb_cluster_data

# Looking at accuracy of all method types in agnes
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(model_data, method = x)$ac
}

map_dbl(m, ac)
```

# creating clusters based on plate discipline #

## Data Manipulation

```{r}
predictors <- c("K_pct", "BB_pct", "OSwing_pct", "ZSwing_pct",
                "OContact_pct", "ZContact_pct", "Swing_pct",
                "Zone_pct", "shadow_rate", "FStrike_pct",
                "Contact_pct", "SwStr_pct", "CStr_pct")

# reference data
fangraphs %>%
  group_by(Season) %>%
  filter(TBF >= 50) %>%
  mutate(across(.cols = predictors, ~ntile(., 100), 
                .names = "{.col}_percentile")) %>%
  ungroup() %>% 
  select(MLBAMID, Season, Role, all_of(predictors),
         contains("percentile")) -> agnes_data 

# model data
agnes_data %>% 
  select(contains("percentile")) -> model_data 
```

## Optimizing K Clusters

```{r}
set.seed(3422354)

# WSS method
fviz_nbclust(model_data, FUN = hcut, method = "wss")$data %>% 
  rename(wss = y) -> wss

# Silhouette method
fviz_nbclust(model_data, FUN = hcut, 
             method = "silhouette")$data %>% 
  rename(mean_silhouette_width = y) -> silhouette

# Gap stat method

clusGap(model_data, FUN = hcut, K.max = 10,
        B = 50, method = "firstSEmax") -> gap_stat

gap_stat[["Tab"]] %>% 
  as_tibble() %>% 
  mutate(clusters = row_number()) -> gap_stat_table

# Visualize gap stat
fviz_gap_stat(gap_stat)

# look at all metrics
merge(
  merge(gap_stat_table, wss, by = "clusters"), 
  silhouette, by = "clusters"
  ) -> all_metrics
```

## Final Clustering

```{r}
set.seed(123)

final <- agnes(model_data, method = "ward", metric = "euclidean")

# optimal pd clusters
k <- 3 
final_clusters <- cutree(as.hclust(final), k = k)
table(final_clusters)

# Visualizing clusters
fviz_cluster(list(data = model_data, 
                  cluster = final_clusters), labelsize = 4)

agnes_data %>% 
  mutate(cluster = final_clusters) -> pd_cluster_data

# Looking at accuracy of all method types in agnes
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(model_data, method = x)$ac
}

map_dbl(m, ac)
```

# creating clusters for reliever roles #

## manipulate pitcher_management data

```{r}
# inherited runners in scoring position
irisp <- c("0-0-1-0", "0-0-1-1", "0-0-1-2",
           "0-1-0-0", "0-1-0-1", "0-1-0-2",
           "0-1-1-0", "0-1-1-1", "0-1-1-2",
           "1-0-1-0", "1-0-1-1", "1-0-1-2",
           "1-1-1-0", "1-1-1-1", "1-1-1-2",
           "1-1-0-0", "1-1-0-1", "1-1-0-2")

pitcher_management %>% 
  mutate(game_situation = as.factor(game_situation),
         bases_empty_no_out = ifelse(game_situation == "0-0-0-0", 1, 0),
         bases_empty_one_out = ifelse(game_situation == "0-0-0-1", 1, 0),
         bases_empty_two_out = ifelse(game_situation == "0-0-0-2", 1, 0),
         rot_no_out = ifelse(game_situation == "0-0-1-0", 1, 0),
         rot_one_out = ifelse(game_situation == "0-0-1-1", 1, 0),
         rot_two_out = ifelse(game_situation == "0-0-1-2", 1, 0),
         ros_no_out = ifelse(game_situation == "0-1-0-0", 1, 0),
         ros_one_out = ifelse(game_situation == "0-1-0-1", 1, 0),
         ros_two_out = ifelse(game_situation == "0-1-0-2", 1, 0),
         rost_no_out = ifelse(game_situation == "0-1-1-0", 1, 0),
         rost_one_out = ifelse(game_situation == "0-1-1-1", 1, 0),
         rost_two_out = ifelse(game_situation == "0-1-1-2", 1, 0),
         rof_no_out = ifelse(game_situation == "1-0-0-0", 1, 0),
         rof_one_out = ifelse(game_situation == "1-0-0-1", 1, 0),
         rof_two_out = ifelse(game_situation == "1-0-0-2", 1, 0),
         corners_no_out = ifelse(game_situation == "1-0-1-0", 1, 0),
         corners_one_out = ifelse(game_situation == "1-0-1-1", 1, 0),
         corners_two_out = ifelse(game_situation == "1-0-1-2", 1, 0),
         rofs_no_out = ifelse(game_situation == "1-1-0-0", 1, 0),
         rofs_one_out = ifelse(game_situation == "1-1-0-1", 1, 0),
         rofs_two_out = ifelse(game_situation == "1-1-0-2", 1, 0),
         loaded_no_out = ifelse(game_situation == "1-1-1-0", 1, 0),
         loaded_one_out = ifelse(game_situation == "1-1-1-1", 1, 0),
         loaded_two_out = ifelse(game_situation == "1-1-1-2", 1, 0),
         irisp_sit = ifelse(game_situation %in% irisp, 1, 0)
         ) %>% 
  group_by(pitcher, game_year) %>% 
  summarise(irisp_pct = sum(irisp_sit)/n(),
            fresh_inning_pct = sum(bases_empty_no_out)/n(),
            n_appearances = n()) -> game_situations

# reference data
merge(game_situations, fangraphs %>% 
        filter(Role == "RP") %>% 
        select(MLBAMID, Season, gmLI, TBF),
      by.x = c("pitcher", "game_year"), by.y = c("MLBAMID", "Season"),
      all.x = T) %>% 
  drop_na() %>% 
  filter(n_appearances >= 15) %>% 
  mutate(bf_per_appearance = TBF / n_appearances) -> pitcher_role_df

predictors <- c("gmLI", "irisp_pct", "fresh_inning_pct",
                "bf_per_appearance")

# reliever role model data
pitcher_role_df %>%
  mutate(across(.cols = predictors, ~ntile(., 100), 
                .names = "{.col}_percentile")) %>% 
  select(contains("percentile")) -> model_data
```

## Optimizing K Clusters

```{r}
set.seed(3422354)

# WSS method
fviz_nbclust(model_data, FUN = hcut, method = "wss")$data %>% 
  rename(wss = y) -> wss

# Silhouette method
fviz_nbclust(model_data, FUN = hcut, 
             method = "silhouette")$data %>% 
  rename(mean_silhouette_width = y) -> silhouette

# Gap stat method

clusGap(model_data, FUN = hcut, K.max = 10,
        B = 50, method = "firstSEmax") -> gap_stat

gap_stat[["Tab"]] %>% 
  as_tibble() %>% 
  mutate(clusters = row_number()) -> gap_stat_table

# Visualize gap stat
fviz_gap_stat(gap_stat)

# look at all metrics
merge(
  merge(gap_stat_table, wss, by = "clusters"), 
  silhouette, by = "clusters"
  ) -> all_metrics
```

## Final Clustering

```{r}
set.seed(123)
final <- agnes(model_data, method = "ward", metric = "euclidean")

# optimal role clusters
k <- 6 
final_clusters <- cutree(as.hclust(final), k = k)
table(final_clusters)

# Visualizing clusters
fviz_cluster(list(data = model_data, 
                  cluster = final_clusters), labelsize = 4)

pitcher_role_df %>% 
  mutate(cluster = final_clusters) -> cluster_data

# Looking at accuracy of all method types in agnes
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(model_data, method = x)$ac
}

map_dbl(m, ac)
```

# merge clusters #

```{r}
merge(pd_cluster_data %>% 
        rename(pd_cluster = cluster) %>% 
        select(MLBAMID, Season, Role, pd_cluster),
      bb_cluster_data %>% 
        rename(bb_cluster = cluster) %>% 
        select(MLBAMID, Season, Role, bb_cluster),
      by = c("MLBAMID", "Season", "Role")) %>% 
  filter(Role == "RP") %>% 
  select(-Role) %>% 
  ungroup() -> clusters

left_join(pitcher_management, cluster_data,
          by = c("pitcher", "game_year")) %>% 
  mutate(role = case_when(cluster == 1 ~ "Closer",
                          cluster == 2 ~ "Middle Reliever",
                          cluster == 3 ~ "Low Leverage Stopper",
                          cluster == 4 ~ "High Leverage Stopper",
                          cluster == 5 ~ "Specialist",
                          cluster == 6 ~ "Long Relief")
         ) -> pitcher_management

left_join(pitcher_management, 
          clusters, 
          by = c("pitcher" = "MLBAMID", "game_year" = "Season")) %>% 
  mutate(pitcher_class = case_when(
    pd_cluster == 1 & bb_cluster == 1 ~ "Extreme GB Junk",
    pd_cluster == 1 & bb_cluster == 2 ~ "GB Junk",
    pd_cluster == 1 & bb_cluster == 3 ~ "Extreme FB Junk",
    pd_cluster == 1 & bb_cluster == 4 ~ "FB Junk",
    pd_cluster == 2 & bb_cluster == 1 ~ "Extreme GB Finesse",
    pd_cluster == 2 & bb_cluster == 2 ~ "GB Finesse",
    pd_cluster == 2 & bb_cluster == 3 ~ "Extreme FB Finesse",
    pd_cluster == 2 & bb_cluster == 4 ~ "FB Finesse",
    pd_cluster == 3 & bb_cluster == 1 ~ "Extreme GB Power",
    pd_cluster == 3 & bb_cluster == 2 ~ "GB Power",
    pd_cluster == 3 & bb_cluster == 3 ~ "Extreme FB Power",
    pd_cluster == 3 & bb_cluster == 4 ~ "FB Power"
  )) -> pitcher_management

# aggregating delta run expectancy by game situation
pitcher_management %>% 
  group_by(pitcher_class, game_situation) %>% 
  summarise(mean_delta_run_exp = mean(delta_run_exp))

# aggregating delta run expectancy by role
pitcher_management %>% 
  group_by(pitcher_class, role_class) %>% 
  summarise(mean_delta_run_exp = mean(delta_run_exp)) %>% 
  drop_na()

# aggregating delta run expectancy by pitcher and season
pitcher_management %>% 
  group_by(pitcher, game_year, role_class, pitcher_class) %>% 
  summarise(mean_delta_run_exp = mean(delta_run_exp),
            n_appearances = n()) %>% 
  drop_na(role_class)
```
